Seeing Double: Indications From {\it TESS} and {\it Gaia} That PTFO$\,$8-8695b is Not a Planet

### 2020-04-16 paper/ms.tex
wrote a sassy introduction

%% Reintroduced the \received and \accepted commands from AASTeX v5.2.
%% Add "Submitted to " argument.
\received{\today}
\revised{---}
\accepted{---}
\submitjournal{AAS journals.}
\shorttitle{Against PTFO$\,$8-8695b}

@@

%\title{Against the Planetary Interpretation of PTFO$\,$8-8695b}
% \title{Seeing Double: Still Against the Planetary Interpretation of PTFO$\,$8-8695b}
% \title{Seeing Double: TESS and Gaia Show That PTFO$\,$8-8695b is Unlikely a
% Planet}
\title{Seeing Double: Indications From {\it TESS} and {\it Gaia} That
PTFO$\,$8-8695b is Not a Planet}

@@

%
% contributing authors: alphabetical
%
% \author[0000-0001-8638-0320]{A. W. Howard}
% \affiliation{Cahill Center for Astrophysics, California Institute of
% Technology, Pasadena, CA 91125, USA}
% %
% \author[0000-0002-2532-2853]{S. B. Howell}
% \affiliation{NASA Ames Research Center, Moffett Field, CA 94035, USA}
% %
% \author[0000-0002-0531-1073]{H. Isaacson}
% \affiliation{Astronomy Department, University of California, Berkeley,
% CA 94720, USA}
% %
% \author{H. Knutson}
% \affiliation{Division of Geological and Planetary Sciences, California
% Institute of Technology, Pasadena, CA 91125, USA}
% %
% \author[0000-0001-7233-7508]{R. A. Matson}
% \affiliation{U.S. Naval Observatory, Washington, DC 20392, USA}
% %

\begin{abstract}
  PTFO$\,$8-8695b is a candidate hot Jupiter in the 7--10 million year old
  Orion-OB1a cluster. It might be the youngest hot Jupiter known.
  We inspected data from TESS and Gaia to clarify whether it is
  actually a planet.
  The TESS lightcurve shows that the dominant variability in this
  system is a sinusoid with a ``long'' period $P_{\rm
  \ell}=11.96\,$hr, previously thought to be a systematic, but in fact likely caused by stellar rotation.  Also present
  is a complex signal, previously identified as the planet candidate,
  that repeats with a ``short'' period $P_{\rm s}= 10.74\,$hr.
  The two signals beat every 4.48 days.
  Although there is a dip in the short-period signal, ground-based
  photometry from the past decade shows that the orbital phase of the
  dip seems to have instantaneously jumped, at least once, and perhaps
  twice.
  Planets do not ``jump'' in orbital phase.
  Furthermore, the Gaia data show that PTFO$\,$8-8695 is probably a
  photometric binary.
  Given the available evidence, we believe that PTFO$\,$8-8695 is a
  binary M dwarf in which one star shows the ``long''
  rotation signal, and the other star is showing ``transient dipping''
  that has also been observed in other young M dwarfs.
  The planetary interpretation seems quite unlikely.
\end{abstract}

%TODO
\keywords{}

@@

\section{Introduction}
If \ptfob\ were a planet, it would be exceptional.
A transiting hot Jupiter, orbiting a $\approx$3$\,$Myr old M dwarf
in ORION would make it the youngest hot Jupiter known.
Its orbital period of only 12 HOURS would also make it the shortest
period hot Jupiter known.

% Section~\ref{sec:observations} of this paper presents all of the
% available transit data as well as the new radial velocity and speckle
% imaging observations.  Section~\ref{sec:analysis} describes our
% analysis of the data, and our interpretation that WASP-4 is being
% pulled around by a brown dwarf or low-mass star.
% Section~\ref{sec:discussion} places this result within the context of
% orbital decay searches, and points out that line-of-sight
% accelerations will be a relatively common type of ``false positive.''
% Section~\ref{sec:conclusions} offers concluding remarks.

@@

\section{TESS Analysis}
\label{sec:analysis}

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f5.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
		{\bf Timing residuals for \ptfob\ from a decade of monitoring.}
		Black points are times of ``dips'', minus the indicated linear
		ephemeris.  The $y$-axis is given in units of phase for the
		short-period signal.  The star shows the binned TESS ephemeris.
		``Dips'' have been observed by \citet{van_eyken_ptf_2012},
		\citet{ciardi_follow-up_2015}, \citet{yu_tests_2015},
		\citet{raetz_yeti_2016}, \citet{onitsuka_multi-color_2017}, and
		\citet{tanimoto_evidence_2020}.  Certain dips ({\it e.g.}, the one
		at phase 0 in mid-2019) are consistent with noise, and were likely
		reported because something was {\it expected}, rather than
		convincingly {\it observed}.  Horizontal dashed lines are drawn at
		$\pm (P_{\rm \ell} - P_{\rm s})/P_{\rm s}$, highlighting what
		could be either a numerical coincidence or an observational bias.
		The orbital phase observed by TESS is consistent with that of
		\citet{tanimoto_evidence_2020}, and quite different from the
		original phase.
		\label{fig:o_minus_c}
	}
\end{figure}

@@

@@

The standard interpretation of sinusoidal modulations for a
pre-main-sequence M dwarf is that we are observing a rotation period.
There is no evidence to suggest that we are seeing anything different.

This is the dominant signal in the system with 10\% amplitude.
Previous analyses, including the discovery study by
\citet{van_eyken_ptf_2012}, saw the same signal ({\it e.g.}, their
Figure~7).  However it was not previously thought to be astrophysical.
\citet{van_eyken_ptf_2012} identified it as a periodogram peak at
$0.9985 \pm 0.0061\,$days, and ascribed it to the observing cadence,
because of its close correspondence to the sidereal day.  While the
TESS data can show significant reflected light from the Earth
\citep[{\it e.g.},][]{luger_tess_2019}, our pixel-level analysis
showed that the signal is specific to only pixels near \ptfo, and no
other pixels.  We therefore conclude that the signal is astrophysical.

>>>>

%% Reintroduced the \received and \accepted commands from AASTeX v5.2.
%% Add "Submitted to " argument.
\received{\today}
\revised{---}
\accepted{---}
\submitjournal{AAS journals.}
\shorttitle{Seeing Double: PTFO$\,$8-8695}

@@

%\title{Against the Planetary Interpretation of PTFO$\,$8-8695b}
% \title{Seeing Double: Still Against the Planetary Interpretation of
% PTFO$\,$8-8695b}
% \title{Seeing Double: TESS and Gaia Show That PTFO$\,$8-8695b is Unlikely a
% Planet}
\title{Seeing Double: Indications From {\it TESS} and {\it Gaia} That
PTFO$\,$8-8695b is Not a Planet}

@@

\begin{abstract}
  PTFO$\,$8-8695b is a candidate hot Jupiter in the 7--10 million year old
  Orion-OB1a cluster. We inspected data from TESS and Gaia to clarify whether
  it is actually a planet.  The TESS lightcurve shows that the dominant
  variability in this system is a sinusoid with a ``long'' period $P_{\rm
  \ell}=11.96\,$hr, likely caused by stellar rotation.  Also present is a
  complex signal, previously identified as the planet candidate, that repeats
  with a ``short'' period $P_{\rm s}= 10.74\,$hr.  The two signals beat every
  4.48 days.  Although there is a dip in the short-period signal, ground-based
  photometry from the past decade shows that the orbital phase of the dip seems
  to have instantaneously jumped, at least once, and perhaps twice.  Planets do
  not ``jump'' in orbital phase.  Furthermore, the Gaia data show that
  PTFO$\,$8-8695 is probably a photometric binary.  Given the evidence, we
  believe that PTFO$\,$8-8695 is a binary M dwarf in which one star shows the
  ``long'' rotation signal, and the other is showing ``transient dipping'' that
  has also been observed in other young M dwarfs.  The planetary interpretation
  seems unlikely.
\end{abstract}

%TODO
\keywords{
	Exoplanet evolution (491),
	Open star clusters (1160),
	Stellar ages (1581),
	Stellar rotation (1629),
	Variable stars (1761)
}

@@

\section{Introduction}
If \ptfob\ were a planet, it would be exceptional.  Transiting a sub-10$\,$Myr
old weak-lined T Tauri M dwarf in Orion, it would be the youngest hot Jupiter
known \citep{van_eyken_ptf_2012}.  Its orbital period of only 10.7 hours would
also give it the shortest period of any known hot Jupiter.  With such a short
period, it would almost certainly have filled its Roche lobe, and would be
actively losing mass to its host star.  Not only that, but the rapidly rotating
host star would also be oblate and gravity darkened, and so the planet's orbit
would likely precess into and out of transitability
\citep{barnes_measurement_2013,ciardi_followup_2015,kamiaka_revisiting_2015}.

Additional lines of evidence could imply further fascinating ``firsts'' for
this candidate planet.  One first would be that its transits are about three
times deeper in optical bandpasses ({\it e.g.,} $g$) than in the near-infrared
({\it e.g.}, $z$-band) \citep{onitsuka_multicolor_2017,tanimoto_evidence_2020}.
A cloud-free hydrogen-dominated planetary atmosphere cannot explain such a
wavelength dependence.  The proposed explanation is that the planet is
surrounded by a dust cloud \citep{tanimoto_evidence_2020}.

A second first could be the detection of direct H$\alpha$ emission from the
planet itself \citep{johnskrull_h_2016}.  While the stellar chromosphere emits
in H$\alpha$, it seems that there is an additional excess H$\alpha$ emission
that could be in phase with the planetary orbit.  The average velocity width of
the excess H$\alpha$ emission is 87$\,$km$\,$s$^{-1}$, and its equivalent width
is 70-80\% that from the stellar chromosphere \citep{johnskrull_h_2016}.  The
proposed explanation is that the outflowing mass from the planet may explain
this excess emission as well \citep{johnskrull_h_2016}.

There are perhaps a few challenges to the planetary interpretation.  They
include that the planet does not seem to emit infrared radiation in occulation,
at least anywhere near the expected amplitude \citep{yu_tests_2015}.  In
addition, despite measurement attempts by multiple investigators, it does not
seem to show the Rossiter effect at the amplitude expected given the rapid
stellar rotation and large planet size
\citep{yu_tests_2015,ciardi_followup_2015}.  Finally, detailed modelling of the
``precession + gravity darkening'' transits has shown that the necessary degree
of gravity darkening is too great, given the spectroscopically observed
equatorial velocity \citep{howarth_reappraisal_2016}.  Additionally, the
gravity-darkened star would need to show photometric variability from its
precession rotation axis, and such variability has not been observed.

While the planetary interpetation has challenges, so too do alternative
explanations.  High-latitude accretion hotspots might produce the observed
H$\alpha$ variability, but \ptfo\ does not have an infrared (IR) excess
associated with the presence of an inner disk.  Low-latitude spots, hot or
cold, struggle to produce the necessary dip durations.  A circumstellar cloud
of dust or gas might be viable, but clumps must be invoked to explain the
short-duration flux dips.  The material might be a remnant of the primordial
gas disk, or it could be ejected and entrained in the stellar magnetosphere.

Section~\ref{sec:observations} presents newly available observations from the
Transiting Exoplanet Survey Satellite (TESS; \citealt{ricker_transiting_2015})
and Gaia \citep{gaia_collaboration_gaia_2018}.  Analyzing the TESS data in
Section~\ref{sec:tess}, we find two distinct signals.  A long sinusoid repeats
every 12.0 hours, and is probably stellar rotation.  A short ``dip + complex
modulation'' repeats every 10.7 hours.  Analyzing the Gaia data in
Section~\ref{sec:gaia}, we show that \ptfo\ is quite likely a photometric
binary.  In Section~\ref{sec:discussion}, we note that the orbital phase of the
dip seems to have instantaneously changed, perhaps twice.  We argue that
PTFO$\,$8-8695 is a binary M dwarf in which one star shows a rotation signal,
and the other is showing ``transient dipping'' that has been observed in other
young M dwarfs, and is likely caused by eclipses of warm particulate clouds.
Section~\ref{sec:conclusions} summarizes our main points.

@@

\section{TESS Analysis}
\label{sec:tess}

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f5.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
		{\bf Timing residuals for \ptfob\ from a decade of monitoring.}
		Black points are times of ``dips'', minus the indicated linear
		ephemeris.  The $y$-axis is given in units of phase for the
		short-period signal.  The star shows the binned TESS ephemeris.
		``Dips'' have been observed by \citet{van_eyken_ptf_2012},
		\citet{ciardi_followup_2015}, \citet{yu_tests_2015},
		\citet{raetz_yeti_2016}, \citet{onitsuka_multi-color_2017}, and
		\citet{tanimoto_evidence_2020}.  Certain dips ({\it e.g.}, the one
		at phase 0 in mid-2019) are consistent with noise, and were likely
		reported because something was {\it expected}, rather than
		convincingly {\it observed}.  Horizontal dashed lines are drawn at
		$\pm (P_{\rm \ell} - P_{\rm s})/P_{\rm s}$, highlighting what
		could be either a numerical coincidence or an observational bias.
		The orbital phase observed by TESS is consistent with that of
		\citet{tanimoto_evidence_2020}, and quite different from the
		original phase.
		\label{fig:o_minus_c}
	}
\end{figure}

@@

%FIXME: summarize existant AO constraints here too
\citet{schmidt_direct_2016}
\citet{lee_evidence_2018}
% \citet{czesla_xray_2019}

@@

The standard interpretation of sinusoidal modulations for a pre-main-sequence M
dwarf is that we are observing a stellar rotation period of 11.96 hours.  This
is the dominant signal in the system with 10\% amplitude, and there is no
evidence to suggest that this signal has any other origin.

The discovery study by \citet{van_eyken_ptf_2012} saw the same signal ({\it
e.g.}, their Figure~7), but identified its alias as a periodogram peak at
$0.9985 \pm 0.0061\,$days. They ascribed it to their observing cadence, because
of its close correspondence to the sidereal day.  While the TESS data can show
significant reflected light from the Earth \citep[{\it
e.g.},][]{luger_tess_2019}, our pixel-level analysis showed that the signal is
specific to only pixels near \ptfo, and no other pixels.  We therefore conclude
that the signal is not a systematic.

We are not the first to reach the conclusion that the long period sinusoidal
modulation is astrophysical.  A follow-up study by
\citet{koen_multicolour_2015} identified the same modes and aliases as
\citet{van_eyken_ptf_2012}, but argued that the $0.50\,{\rm d}$ signal was
astrophysical.  Using archival photometry and photometry from the YETI global
telescope network, \citet{raetz_yeti_2016} eventually came to the conclusion
that the that the $0.50\,{\rm d}$ signal was indeed from stellar rotation.  The
TESS data strongly support this conclusion.
